---
title: "Download and preprocess the supplemental tables in Yaron-Barir et al. 2024 (doi:10.1038/s41586-024-07407-y)"
author: "Florian Geier"
date: "16.September.2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 100)
```

```{r setup}
library(here) # set relative locations

library(readxl) # read raw data (excel files)
library(dplyr)
library(tidyr)
library(stringr)

library(Homo.sapiens) # add kinase gene annotations
```

## Download raw data

```{r download_data}
here::i_am('inst/scripts/make-data-Y.Rmd')

download.file(url='https://static-content.springer.com/esm/art%3A10.1038%2Fs41586-024-07407-y/MediaObjects/41586_2024_7407_MOESM3_ESM.xlsx',
              destfile=here::here('inst','extdata','41586_2024_7407_MOESM3_ESM.xlsx'))

download.file(url='https://static-content.springer.com/esm/art%3A10.1038%2Fs41586-024-07407-y/MediaObjects/41586_2024_7407_MOESM4_ESM.xlsx', 
              destfile=here::here('inst','extdata','41586_2024_7407_MOESM4_ESM.xlsx'))

download.file(url='https://static-content.springer.com/esm/art%3A10.1038%2Fs41586-024-07407-y/MediaObjects/41586_2024_7407_MOESM5_ESM.xlsx',
              destfile=here::here('inst','extdata','41586_2024_7407_MOESM5_ESM.xlsx'))
```

## Kinase annotation

- Extract annotation for each kinase matrix and add Entrez ID and gene descriptions from Bioc annotation databases.

- Fix a couple of wrong gene symbols by hand

```{r kinase_annotation}
anno <- readxl::read_excel(here::here('inst','extdata','41586_2024_7407_MOESM3_ESM.xlsx'), 
                           sheet='Table_S1_Data')
anno <- anno |>
  dplyr::filter(Species %in% c('human','Tribolium castaneum') & !is.na(UNIPROT_ID)) |> # kinase PWM available 
  dplyr::mutate(MatrixName = MATRIX_NAME,
                UniprotID = UNIPROT_ID,
                GeneName = GENENAME,
                KinaseSubType = SUBTYPE,
                KinaseFamily = FAMILY) |>
  dplyr::select(MatrixName, GeneName, UniprotID, KinaseSubType, KinaseFamily) 

## correct by hand wrong entries
anno$GeneName[anno$GeneName == "IRR"] <- "INSRR" # insulin receptor related receptor
anno$MatrixName[anno$GeneName == "MST1R"] <- "MST1R" ## wrong MatrixName (RON)
anno$GeneName[anno$GeneName == "MKK4"] <- "MAP2K4" # mitogen-activated protein kinase kinase 
anno$GeneName[anno$GeneName == "MKK6"] <- "MAP2K6"
anno$GeneName[anno$GeneName == "MKK7"] <- "MAP2K7"
anno$GeneName[anno$GeneName == "PDHK1"] <- "PDK1" # pyruvate dehydrogenase kinase 
anno$GeneName[anno$GeneName == "PDHK3"] <- "PDK3"
anno$GeneName[anno$GeneName == "PDHK4"] <- "PDK4"

## add '_TYR' to non-canonical TK MatrixName
anno <- anno |> 
    dplyr::mutate(MatrixName = ifelse(KinaseSubType == 'ncTK', paste0(MatrixName, '_TYR'), MatrixName))  

## add EntrezID and gene description
sel <- AnnotationDbi::select(Homo.sapiens, 
                             keys = anno |> dplyr::pull(GeneName), 
                             keytype = "SYMBOL", 
                             columns = c("GENEID","GENENAME"))
## remove wrong gene for symbol TEC
sel <- sel[!sel$GENEID == "100124696",] 

df <- anno |> 
  dplyr::left_join(sel, by = join_by(GeneName == SYMBOL)) |>
  dplyr::rename(Description = GENENAME, EntrezID = GENEID) |>
  dplyr::relocate(EntrezID, Description, .after = UniprotID)

write.csv(df, 
          file = here::here('inst','extdata','TyrosineKinaseAnnotation.csv'), 
          row.names = FALSE)
```

## Process tyrosine kinase matrices

- PWMs are taken from `tyrosine_all_norm_scaled_matric` sheet.

- The PWMs contain lower case letters for phosphorylated, non-central S/T/Y residues.

```{r process_pwm}
data_scaled <- readxl::read_excel(here::here('inst','extdata','41586_2024_7407_MOESM4_ESM.xlsx'),
                                  sheet='tyrosine_all_norm_scaled_matric') |>
  dplyr::rename(Matrix = 1) |> 
  tidyr::pivot_longer(cols = -Matrix,
                      names_to = "Position",
                      values_to = "Score") |>
  tidyr::separate_wider_regex(cols = "Position", 
                              patterns = c(Position = "[^A-z]+", 
                                           AA = "[A-z]+")) |>
  dplyr::mutate(Position = as.integer(Position)) # 

PWM <- data_scaled |> 
  dplyr::mutate(Score = log2(Score)) |>
  dplyr::arrange(Matrix, Position, AA)

write.csv(PWM, 
          file = gzfile(here::here('inst','extdata','TyrosineKinasePWM.csv.gz')), 
          row.names = FALSE)
```

## PWM background scores

- Background peptide sequences are taken from the `SITE_+/-7_AA`
  column of the corresponding excel sheet, and the central 11 AA are
  extracted (central phosphosite at position 6)

- Next, all kinase PWMs defined above are matched against all unique
  background peptides

- The resulting score vectors are combine into a data frame and saved.

```{r background_scores}
background <- readxl::read_excel(here::here('inst','extdata',"41586_2024_7407_MOESM5_ESM.xlsx"),
                                 sheet=2)

sites <- background |>
  dplyr::mutate(sites = stringr::str_sub(background$`SITE_+/-7_AA`, start = 3, end = 13)) |>
  dplyr::pull(sites) |>
  unique()

length(sites) # reported in paper: 7,315 but is 6659 unique sites in the table

source(here::here('R','score_phosphosites.R')) # functionality to score sites

pwms <- read.csv(here::here('inst','extdata','TyrosineKinasePWM.csv.gz'))
PWM <- lapply(split(pwms, pwms$Matrix), function(x) {
    x <- x |> 
        dplyr::select(-Matrix) |> 
        tidyr::pivot_wider(values_from=Score, 
                           names_from=Position) |>
        dplyr::mutate(`0` = NA_real_, .before=`1`)
    y <- as.matrix(x |> dplyr::select(-AA))
    rownames(y) <- x |> dplyr::pull(AA)
    y
})
    
scores <- scorePhosphosites(PWM, sites, scoreType = 'lod')
  
write.csv(data.frame(PhosphoSite = sites, scores), 
          file = gzfile(here::here('inst','extdata','TyrosineKinaseBackgroundScores.csv.gz')),
          row.names = FALSE)

## compute quantiles
probs <- seq(0,1,length.out=1001)
quants <- sapply(colnames(scores), function(x)  quantile(scores[,x], probs))
write.csv(data.frame(Quantiles = probs, quants), 
          file = gzfile(here::here('inst','extdata','TyrosineKinaseBackgroundQuantiles.csv.gz')), 
          row.names = FALSE)
```

## Session information

```{r session_info}
sessionInfo()
```
